---
title: 'Practical 6: Software and GLMs'
author: "VIBASS"
date: "July 2023"
output:
  html_vignette:
    fig_caption: yes
    number_sections: yes
    toc: yes
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Practical 6: Software and GLMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

# Software for Bayesian Statistical Analysis

So far, only simple Bayesian models with conjugate priors have been considered.
As explained in previous practicals, when the posterior distribution is not
available in closed form, MCMC algorithms such as the Metropolis-Hastings or
Gibbs Sampling can be used to obtain samples from it.

In general, posterior distributions are seldom available in closed form and
implementing MCMC algorithms for complex models can be technically difficult
and very time-consuming.

For this reason, in this Practical we start by looking at a number of `R`
packages to fit Bayesian statistical models. These packages will equip us
with tools which can be used to deal with more complex models efficiently,
without us having to do a lot of extra coding ourselves. Fitting Bayesian
models in `R` will then be much like fitting non-Bayesian models, using
model-fitting functions at the command line, and using standard syntax for
model specification.

## BayesX and INLA

In particular, the following two software packages will be considered:

* `BayesX`

* `INLA`

`BayesX` (http://www.bayesx.org/) implements MCMC methods to obtain samples
from the joint posterior and is conveniently accessed from R via the package
`R2BayesX`.

`INLA` (https://www.r-inla.org/) is based on producing (accurate)
approximations to the marginal distributions of the model parameters.
Although this can be enough most of the time, making multivariate inference
with `INLA` can be difficult or impossible. However, in many cases this is
not needed and `INLA` can fit some classes of models in a fraction of the
time it takes with MCMC.

Both `R2BayesX` and `INLA` have a very simple interface to define models
using a `formula` (in the same way as with `glm` and `gam`).

While `R2BayesX` can be installed from CRAN, `INLA` is not on CRAN and
needs to be installed from a specific repository.

## Other Bayesian Software

NEEDS WRITING

# Bayesian Logistic Regression

## Model Formulation
[RECAP OF LOGISTIC MODEL FROM SLIDES]

## Example: Fake News

The `fake_news` data set in the `bayesrules` package in `R` contains
information about 150 news articles, some real news and some fake news.

In this example, we will look at trying to predict whether an article of news
is fake or not given three explanatory variables.

We can use the following code to extract the variables we want from the
data set without having to load the package itself:

```{r}
fakenews <- bayesrules::fake_news[,c("type","title_has_excl","title_words","negative")]
```

The response variable `type` takes values `fake` or `real`, which should be
self-explanatory. The three explanatory variables are:

* `title_has_excl`, whether or not the article contains an excalamation mark (values `TRUE` or `FALSE`);

* `title_words`, the number of words in the title (a positive integer); and

* `negative`, a sentiment rating, recorded on a continuous scale.




## Exercises

# Bayesian Poisson Regression

## Model Formulation
[RECAP OF POISSON MODEL FROM SLIDES]

## Example: Emergency Room Complaints

Example: in faraway package
esdcomp: Complaints about emergency room doctors

Data was recorded on 44 doctors working in an emergency service at a
hospital to study the factors affecting the number of complaints received.

## Exercises



